Increased perceptions of autonomy through choice fail to enhance motor skill retention

### 2021-12-28 docs/accepted/accepted_manuscript.tex
init

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  doc, donotrepeattitle,floatsintext]{apa7}
\title{Increased perceptions of autonomy through choice over one's observation schedule fail to enhance motor skill retention}
\author{Laura St.~Germain\textsuperscript{1}, Allison Williams\textsuperscript{1}, Noura Balbaa\textsuperscript{1}, Andrew Poskus\textsuperscript{1}, Olena Leshchyshen\textsuperscript{1}, Keith R. Lohse\textsuperscript{2}, \& Michael J. Carter\textsuperscript{1}}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Increased perceptions of autonomy through choice over one's observation schedule fail to enhance motor skill retention},
  pdfauthor={Laura St.~Germain1, Allison Williams1, Noura Balbaa1, Andrew Poskus1, Olena Leshchyshen1, Keith R. Lohse2, \& Michael J. Carter1},
  pdflang={en-EN},
  pdfkeywords={Motor learning; Pre-registered; Self-controlled; Observation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Motor learning; Pre-registered; Self-controlled; Observation}
\usepackage{csquotes}
\usepackage{pdflscape}
\raggedbottom
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

PsyArXiv preprint: \url{https://doi.org/10.31234/osf.io/s34fg} \vspace{1em}

@@

\abstract{
There has been growing research interest in the effects that motivation plays in motor learning, and specifically how autonomy, competence, and social relatedness may directly benefit the learning process. Here, we present a preregistered manipulation of autonomy-support by providing learners with choice during the practice of a speed cup-stacking task. One group was given control over when a video demonstration was provided and the viewing speed. A yoked control group received an identical demonstration schedule, but without choice (as their schedule was matched to a participant with choice). Critically, we addressed a gap in the literature by adding a yoked group who was explicitly told that they were being denied choice and that their schedule was chosen by another participant. We found no statistically significant learning differences between groups, despite finding evidence that providing choice increased perceived autonomy. Equivalence tests further showed that although the groups were not statistically equivalent, the effect size is likely too small to practically study the effects of autonomy-support through choice in most motor learning labs. These findings add to a growing body of research that questions a causal role of autonomy-support on motor learning, and the robustness of the so-called self-controlled learning advantage.
}

@@

\vspace{-0.5em}

\noindent

\begin{center}
\textcolor{blue}{Accepted for publication in \emph{Journal of Experimental Psychology: Human Perception \& Performance}}
\end{center}

\vspace{0.5em}

@@

\caption{\label{tab:table1}Group specific components from the instructions detailing the observation schedule that would be experienced during Acquisition.}
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{8em}>{\raggedright\arraybackslash}p{33em}}
\toprule
  & Acquisition instructions\\
\midrule
Self-Controlled & Before each trial, you will be asked whether you wish to watch a modeled demonstration of the task. If you choose YES, you will then be asked whether you want to watch the video in real-time or in slow-motion.\\
\addlinespace
Traditional Yoked & Before each trial, you may or may not watch a modeled demonstration of the task based on a pre-determined schedule. If you observe a model, it might be presented in real-time or in slow-motion based on the pre-determined schedule.\\
\addlinespace
Explicit Yoked & Before each trial, you may or may not watch a modeled demonstration of the task based on the schedule another participant selected. If you observe a model, it might be presented in real-time or in slow-motion based on what that participant selected.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:table2}Cronbach's alpha for each questionnaire at each time point.}
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{7em}}
\toprule
  & After pre-test & After trial 5 & After trial 25 & Before retention\\
\midrule
Perceived autonomy & 0.62 & 0.79 & 0.79 & 0.84\\
\addlinespace
Perceived competence & 0.90 & 0.92 & 0.94 & 0.92\\
\addlinespace
Intrinsic motivation & 0.86 & 0.90 & 0.92 & 0.92\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\caption{\textbf{Physical performance data of the experiment.} The Self-Controlled group is shown in red with solid lines and/or circles, the Traditional Yoked group is shown in blue with dotted and/or triangles, and the Explicit Yoked group is shown in orange with dashed and/or squares. \textbf{(A).} Trial-by-trial stacking time (s) data averaged across participants within each group (shaded area represents standard error). Splits between line segments denote the end of one phase and the start of the next. Pre-test (Trials 1 to 5) and Acquisition (Trials 6 to 30) occurred on Day 1 and Retention (Trials 31 to 35) occurred on Day 2, approximately 24-hours later. \textbf{(B).} Mean stacking time (s) for each group was computed by averaging the data from A into 7 blocks of trials. Error bars represent 95\% confidence intervals. Splits between line segments denote the end of one phase and the start of the next. \textbf{(C).} Mean stacking time (s) adjusted for pre-test for each group. Error bars represent 95\% confidence intervals. \textbf{(D).} Scatterplot showing the relationship between each participant's mean stacking time (s) in pre-test and retention.}\label{fig:fig1}
\end{figure}

\caption{\label{tab:table3}The mean frequencies for each Acquisition block of requesting to watch the video demonstration and of those requests, the mean frequency of choosing the regular viewing speed.}
\fontsize{10}{12}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{5em}l}
\toprule
  & Block 1 & Block 2 & Block 3 & Block 4 & Block 5\\
\midrule
Model viewing frequency & 32.6\% & 16.9\% & 12.0\%\% & 12.9\% & 9.3\%\\
\addlinespace
Regular speed frequency & 33.8\% & 38.1\% & 50.0\% & 34.4\% & 30.4\%\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\textbf{Questionnaire data.} Groups means for perceived autonomy \textbf{(A)}, intrinsic motivation \textbf{(B)}, and perceived competence \textbf{(C)} after pre-test, trial 5, and trial 25 on Day 1, and before retention on Day 2. The Self-Controlled group is shown in red circles, the Traditional Yoked group is shown in blue triangles, and the Explicit Yoked group is shown in orange squares. Dots represent individual participants in each group. Scores could range from 1 (Strongly disagree) to 7 (Strongly agree). Error bars represent 95\% confidence intervals.}\label{fig:fig2}
\end{figure}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man, donotrepeattitle,floatsintext]{apa7}
\title{Increased perceptions of autonomy through choice fail to enhance motor skill retention}
\author{Laura St.~Germain\textsuperscript{1}, Allison Williams\textsuperscript{1}, Noura Balbaa\textsuperscript{1}, Andrew Poskus\textsuperscript{1}, Olena Leshchyshen\textsuperscript{1}, Keith R. Lohse\textsuperscript{2}, \& Michael J. Carter\textsuperscript{1}}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Increased perceptions of autonomy through choice fail to enhance motor skill retention},
  pdfauthor={Laura St.~Germain1, Allison Williams1, Noura Balbaa1, Andrew Poskus1, Olena Leshchyshen1, Keith R. Lohse2, \& Michael J. Carter1},
  pdflang={en-EN},
  pdfkeywords={Motor learning; Pre-registered; Self-controlled; Observation},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

@@

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{Motor learning; Pre-registered; Self-controlled; Observation\newline\indent Word count: 4585}
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\usepackage{pdflscape}
\pagewiselinenumbers
\raggedbottom
\ifXeTeX
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[main=english]{babel}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\fi
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

@@

PsyArXiv preprint: \url{https://doi.org/10.31234/osf.io/s34fg}

@@

\abstract{
There has been growing research interest in the effects that motivation plays in motor learning, and specifically how autonomy, competence, and social relatedness may directly benefit the learning process. Here, we present a preregistered manipulation of autonomy-support by providing learners with choice during the practice of a speed cup-stacking task. One group was given control over when a video demonstration was provided and the viewing speed. A yoked control group received an identical demonstration schedule, but without choice (as their schedule was matched to a participant with choice). Critically, we addressed a gap in the literature by adding a yoked group who was explicitly told that they were being denied choice and that their schedule was chosen by another participant. We found no statistically significant learning differences between groups, despite finding evidence that providing choice increased perceived autonomy. Equivalence tests further showed that although the groups were not statistically equivalent, the effect size is likely too small to practically study the effects of autonomy-support through choice in most motor learning labs. These findings add to a growing body of research that questions a causal role of autonomy-support on motor learning, and the robustness of the so-called self-controlled learning advantage.

\noindent \textbf{\emph{Public Significance Statement}}\\
The present study suggests that although university-aged students may experience a boost in feelings of autonomy when given choice over features of their learning environment, this boost may not enhance learning of a new motor skill. Our results also highlight the need for higher powered and pre-registered motor learning experiments if coaches and practictioners are to be provided with reliable recommendations for training and rehabilitation protocols.
}

@@

@@

\caption{\label{tab:table1}Group specific components from the instructions detailing the observation schedule that would be experienced during Acquisition.}
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{8em}>{\raggedright\arraybackslash}p{33em}}
\toprule
  & Acquisition instructions\\
\midrule
Self-Controlled & Before each trial, you will be asked whether you wish to watch a modeled demonstration of the task. If you choose YES, you will then be asked whether you want to watch the video in real-time or in slow-motion.\\
\addlinespace
Traditional Yoked & Before each trial, you may or may not watch a modeled demonstration of the task based on a pre-determined schedule. If you observe a model, it might be presented in real-time or in slow-motion based on the pre-determined schedule.\\
\addlinespace
Explicit Yoked & Before each trial, you may or may not watch a modeled demonstration of the task based on the schedule another participant selected. If you observe a model, it might be presented in real-time or in slow-motion based on what that participant selected.\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{\label{tab:table2}Cronbach's alpha for each questionnaire at each time point.}
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{10em}>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{6em}>{\raggedright\arraybackslash}p{7em}}
\toprule
  & After pre-test & After trial 5 & After trial 25 & Before retention\\
\midrule
Perceived autonomy & 0.62 & 0.79 & 0.79 & 0.84\\
\addlinespace
Perceived competence & 0.90 & 0.92 & 0.94 & 0.92\\
\addlinespace
Intrinsic motivation & 0.86 & 0.90 & 0.92 & 0.92\\
\bottomrule
\end{tabular}
\end{table}

@@

\caption{Physical performance data of the experiment. The Self-Controlled group is shown in red with solid lines and/or circles, the Traditional Yoked group is shown in blue with dotted and/or triangles, and the Explicit Yoked group is shown in orange with dashed and/or squares. \textbf{A}. Trial-by-trial stacking time (s) data averaged across participants within each group (shaded area represents standard error). Splits between line segments denote the end of one phase and the start of the next. Pre-test (Trials 1 to 5) and Acquisition (Trials 6 to 30) occurred on Day 1 and Retention (Trials 31 to 35) occurred on Day 2, approximately 24-hours later. \textbf{B}. Mean stacking time (s) for each group was computed by averaging the data from \textbf{A} into 7 blocks of trials. Error bars represent 95\% confidence intervals. Splits between line segments denote the end of one phase and the start of the next. \textbf{C}. Mean stacking time (s) adjusted for pre-test for each group. Error bars represent 95\% confidence intervals. \textbf{D}. Scatterplot showing the relationship between each participant's mean stacking time (s) in pre-test and retention.}\label{fig:fig1}
\end{figure}

\caption{\label{tab:table3}The mean frequencies for each Acquisition block of requesting to watch the video demonstration and of those requests, the mean frequency of choosing the regular viewing speed.}
\fontsize{11}{13}\selectfont
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{13em}>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{5em}>{\raggedright\arraybackslash}p{5em}l}
\toprule
  & Block 1 & Block 2 & Block 3 & Block 4 & Block 5\\
\midrule
Model viewing frequency & 32.6\% & 16.9\% & 12.0\%\% & 12.9\% & 9.3\%\\
\addlinespace
Regular speed frequency & 33.8\% & 38.1\% & 50.0\% & 34.4\% & 30.4\%\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

\caption{Questionnaire data. Groups means for \textbf{A} perceived autonomy, \textbf{B} intrinsic motivation, and \textbf{C} perceived competence after pre-test, trial 5, and trial 25 on Day 1, and before retention on Day 2. The Self-Controlled group is shown in red circles, the Traditional Yoked group is shown in blue triangles, and the Explicit Yoked group is shown in orange squares. Dots represent individual participants in each group. Scores could range from 1 (Strongly disagree) to 7 (Strongly agree). Error bars represent 95\% confidence intervals.}\label{fig:fig2}
\end{figure}
